---
title: "Project Assignment"
author: "Stephen Sani"
date: "June 24, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##Overview
The aim of the project is to predict the manner in which 6 participants performed some exercise as described below. This is the "classe" variable in the training set.

## Data Loading and Exploratory Analysis
a) Dataset Overview
The training data for this project are available on:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available on:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
The data for this project come from http://groupware.les.inf.puc-rio.br/har. 

### Environment Preparation
We first upload the R libraries that are necessary for the complete analysis.

```{r}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
library(corrplot)
set.seed(12345)
```


### Data Loading and Cleaning
The next step is loading the dataset from the URL provided above. The training dataset is then partinioned in 2 to create a Training set (70% of the data) for the modeling process and a Test set (with the remaining 30%) for the validations. The testing dataset is not changed and will only be used for the quiz results generation.

### set the URL for the download
```{r}
UrlTrain <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
UrlTest <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
```


### download and read the datasets
```{r}
training <- read.csv(url(UrlTrain))
testing <- read.csv(url(UrlTest))
```

# create a partition with the training dataset
```{r}
inTrain <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)
```


Both created datasets have 160 variables. Those variables have plenty of NA, that can be removed with the cleaning procedures below. The Near Zero variance (NZV) variables are also removed and the ID variables as well.

### remove variables with Nearly Zero Variance
```{r}
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet <- TestSet[, -NZV]
dim(TrainSet)
dim(TestSet)
```


### remove variables that are mostly NA
```{r}
AllNA <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet <- TestSet[, AllNA==FALSE]
dim(TrainSet)
dim(TestSet)
```


# remove identification only variables (columns 1 to 5)
```{r}
TrainSet <- TrainSet[, -(1:5)]
TestSet <- TestSet[, -(1:5)]
dim(TrainSet)
dim(TestSet)
```


With the cleaning process above, the number of variables for the analysis has been reduced to 54 only.

## Correlation Analysis
A correlation among variables is analysed before proceeding to the modeling procedures.
```{r}
corMatrix <- cor(TrainSet[, -54])
corrplot(corMatrix, order = "FPC", method = "color", type = "lower",
tl.cex = 0.8, tl.col = rgb(0, 0, 0))
```

The highly correlated variables are shown in dark colors in the graph above. To make an evem more compact analysis, a PCA (Principal Components Analysis) could be performed as pre-processing step to the datasets. Nevertheless, as the correlations are quite few, this step will not be applied for this assignment.

## Prediction Model Building
Three methods will be applied to model the regressions (in the Train dataset) and the best one (with higher accuracy when applied to the Test dataset) will be used for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model, as described below. A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

### Method 1: Random Forest
Model fit
```{r}
set.seed(12345)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf",
trControl=controlRF)
modFitRandForest$finalModel
```


Prediction on Test dataset
```{r}
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest <- confusionMatrix(predictRandForest, TestSet$classe)
confMatRandForest
```


plot matrix results
```{r}
plot(confMatRandForest$table, col = confMatRandForest$byClass,
main = paste("Random Forest - Accuracy =",
round(confMatRandForest$overall['Accuracy'], 4)))
```


### Method 2: Decision Trees
model fit
```{r}
set.seed(12345)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
fancyRpartPlot(modFitDecTree)
```


prediction on Test dataset
```{r}
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, TestSet$classe)
confMatDecTree
```


plot matrix results
plot(confMatDecTree$table, col = confMatDecTree$byClass,
main = paste("Decision Tree - Accuracy =",
round(confMatDecTree$overall['Accuracy'], 4)))

### Method 3: Generalized Boosted Model
model fit
```{r}
set.seed(12345)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM <- train(classe ~ ., data=TrainSet, method = "gbm",
trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```


prediction on Test dataset
```{r}
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, TestSet$classe)
confMatGBM
```


plot matrix results
```{r}
plot(confMatGBM$table, col = confMatGBM$byClass,
main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))

```

## Applying the Selected Model to the Test Data
The accuracy of the 3 regression modeling methods above are:
a. Random Forest : 0.9963
b. Decision Tree : 0.7368
c. GBM : 0.9839
In that case, the Random Forest model will be applied to predict the 20 quiz results (testing dataset) as shown below.
```{r}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```
## Credits
The data used in this analysis is provided by http://groupware.les.inf.puc-rio.br/har .

